---
title: "Project 4"
author: "David Nguyen, Huy Ngo"
date: "March 23, 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Plinko

```{r message = FALSE, warning=FALSE}
library(dplyr); library(tidyr); library(ggplot2)
library(nnet)
library(MASS)
```

1)	Plinko is a game played on The Price is Right TV show. Contestants drop a circular disk, referred to as a “chip”, from the top of a board. The chip travels down the board and bumps into pegs that can change its direction. Eventually, the chip will reach the bottom of the board and land in a slot. Each slot has a dollar amount that the contestant wins. 
A few years ago a graduate student watched all Plinko airings on The Price is Right over one year. She collected information about where the chip was dropped and where the chip landed. This data is available in the plinko.csv file on the graded materials web page of the course website. Below is the data:

```{r}
plinko <- read.csv("data/plinko.csv") 
plinko %>% knitr::kable()
```

where 

* drop: Slot the chip was dropped above at the top of the board with 1 = $100, 2 = $500, 3 = $1,000, 4 = $0, and 5 = $10,000
* win: Slot the chip landed in at the bottom of the board with 1 = $100, 2 = $500, 3 = $1,000, 4 = $0, and 5 = $10,000
* count: Frequency of occurrences during the year

a)	(2 points) Convert the win variable to be of a factor type. Show verification that this variable is a factor. 

```{r}
plinko$win <- factor(plinko$win)
class(plinko$win)
```

b)	For a multinomial regression model using win as the response, complete the parts below. Use winning $100 (slot 1) as the baseline category.  


    i) (3 points) Estimate and state the model using drop as one linear term. 

```{r}
mod.fit1 <- multinom(win ~ drop, data = plinko, weights = count)
summary(mod.fit1)
```

The estimated model is:

\begin{align*}
\log(\hat \pi_2 / \hat \pi_1) & = 0.619 + 0.044 \text{drop} \\
\log(\hat \pi_3 / \hat \pi_1) & = 0.133 + 0.332 \text{drop} \\
\log(\hat \pi_4 / \hat \pi_1) & = -0.454 + 0.449 \text{drop} \\
\log(\hat \pi_5 / \hat \pi_1) & = -1.322 + 0.467 \text{drop} \\
\end{align*}

ii) (3 points) Estimate the probabilities of winning each possible dollar amount for each drop location. 

```{r}
pred1 <- data.frame(drop = 1:5)
pred1 <- cbind(pred1, predict(mod.fit1, 
                     newdata = data.frame(drop = 1:5), 
                     type = "prob"))
pihat_labels <- c("drop", "$\\hat \\pi_1$", "$\\hat\\pi_2$", "$\\hat\\pi_3$",
                  "$\\hat\\pi_4$", "$\\hat\\pi_5$")
pred1 %>% knitr::kable(digits = 3, 
                       col.names = pihat_labels, escape = FALSE)
```


iii) (5 points) Plot the estimated model. Make sure to include a legend to differentiate among the different lines on the plot. Interpret the plot. 

```{r}
pred1 %>% pivot_longer(cols = 2:6, names_to = "win", values_to = "probability") %>%
  ggplot() +
  geom_line(aes(x = drop, y = probability, col = win)) +
  labs(title = "Estimated probability for each win type",
       y = expression(hat(pi[j]))) +
  theme_bw()
```

The probabilities of winning \$100 or \$500 is a decreasing function of drop position whereas the probabilities of winning $0, $1,000, or $10,000 increase as drop position is increased, i.e., the drop occurs closer to the center of the board.
The probability of winning $10,000 is highest directly when dropped directly above that position (drop = 5), but this is also where the probability of winning nothing is highest.

iv)	(5 points) Estimate odds ratios corresponding to the drop variable and calculate the corresponding 95% confidence intervals. Interpret the intervals. Use winning $100 compared to winning another amount in your odds ratios.

```{r}
# get vector of beta_j1  for drop j = 2, ..., 5
beta.hat.1 <- coef(mod.fit1)[,2]
# get OR with respect to drop = 1 ($100) 
# per 1 unit increment closer to the center
OR.hat.1 <- exp(beta.hat.1)
# confidence intervals
# array with index [par = c(interecept, drop), CI = (lower, upper), drop = j = 2:5]
CI.beta1 <- confint(mod.fit1, level = 0.95)
# grab CIs for beta_j1
# columns are lower, upper
CI.OR.beta1 <- t(rbind(CI.beta1[2, ,])) 
CI.OR.beta1 <- exp(CI.OR.beta1) # transorm to OR scale
# combine OR point and interval estimates
OR.hat.table1 <- data.frame(drop = 2:5, OR.hat.1, CI.OR.beta1)
ORhat_labels <- c("drop", "$\\hat{OR}$", "lower 95% CI", "upper 95% CI")
OR.hat.table1 %>% knitr::kable(digits = 2, row.names = FALSE,
                               col.names = ORhat_labels, escape = FALSE)
```

* The estimated odds of winning $500 vs. $100 is 1.04 times larger for every 1 unit increment closer to the center of the board. We are 95% confident that the odds of winning $500 vs. $100 is between 0.50 and 2.16 times larger for for every 1 unit increment closer to the center of the board. Since the confidence interval includes 1, we would fail to reject the null hypothesis that drop position has no effect of the odds of winning $500 vs. $100.
* The estimated odds of winning $1,000 vs. $100 is 1.39 times larger for every 1 unit increment closer to the center of the board. We are 95% confident that the odds of winning $1,000 vs. $100 is between 0.70 and 2.77 times larger for for every 1 unit increment closer to the center of the board. Since the confidence interval includes 1, we would fail to reject the null hypothesis that drop position has no effect of the odds of winning $1,000 vs. $100.
* The estimated odds of winning $0 vs. $100 is 1.57 times larger for every 1 unit increment closer to the center of the board. We are 95% confident that the odds of winning $0 vs. $100 is between 0.78 and 3.16 times larger for for every 1 unit increment closer to the center of the board. Since the confidence interval includes 1, we would fail to reject the null hypothesis that drop position has no effect of the odds of winning $0 vs. $100.
* The estimated odds of winning $10,000 vs. $100 is 1.60 times larger for every 1 unit increment closer to the center of the board. We are 95% confident that the odds of winning $10,000 vs. $100 is between 0.71 and 3.57 times larger for for every 1 unit increment closer to the center of the board. Since the confidence interval includes 1, we would fail to reject the null hypothesis that drop position has no effect of the odds of winning $10,000 vs. $100.

v) (3 points) Perform a LRT to evaluate the importance of the drop variable. Make sure to include the hypotheses in the correct symbolic form. 

$H_0: \beta_{j1} = 0$ for $j = 2, \ldots, 5$

$H_A: \text{at least one } \beta_{j1} \neq 0$

```{r}
car::Anova(mod.fit1, test = "LR")
```

There is not statistical evidence that the drop position has an effect on the log-odds of winning 0, 500, 1000, or 10,000 dollars compared to winning \$100 ($\Lambda = 3.5055, p = 0.477$). This is consistent with our previous finding that all the 95% confidence intervals for the OR included 1.

c)	(3 points) Contestants are often focused on winning \$10,000 (slot 5) when they play the game. For this reason, it may be preferred to use this amount won as the baseline category. What is the estimated model using \$10,000 as the baseline category? 

```{r}
plinko$win_reordered <- factor(plinko$win, levels = c(5,1,2,3,4))
mod.fit_10000 <- multinom(win_reordered ~ drop, data = plinko, weights = count)
summary(mod.fit_10000)
```

The estimated model is:

\begin{align*}
\log(\hat \pi_1 / \hat \pi_5) & = 1.3224033 -0.4671048 \text{drop} \\
\log(\hat \pi_2 / \hat \pi_5) & = 1.9422748 -0.4235397 \text{drop} \\
\log(\hat \pi_3 / \hat \pi_5) & = 1.4553465 -0.1347888 \text{drop} \\
\log(\hat \pi_4 / \hat \pi_5) & = 0.8684506 -0.0180170 \text{drop} \\
\end{align*}

d)	(3 points) Why would a proportional odds regression model potentially be better than a multinomial regression model for this data? 

The win response categories could logically be ordered by the size of the reward. A proportional odds regression model may be better than a multinomial model because if we reject the null hypothesis, then we can state that there is evidence for an increasing (decreasing) trend in the log-odds for $\beta_1 > 0$ ($\beta_1 < 0$). This allows us to test a more specific hypothesis: does dropping the coin towards the center of the board increase the log odds of earning more money?

e)	For a proportional odds regression model using win as the response, complete the parts below. Use the response ordering given by its numerical values (“1” is the lowest value and “5” is the highest value). 

    i)	(3 points) Estimate and state the model using drop as one linear term.  

```{r}
mod.fit2 <- polr(win ~ drop, data = plinko, weights = count, method = "logistic")
summary(mod.fit2)
```

$$logit(P(Y \leq j)) = \beta_{j0} - 0.2923 \times \text{drop}$$

where $\beta_{10} = -1.3169$, $\beta_{20} = 0.0770$, $\beta_{30} = 1.4796$, $\beta_{40} = 3.0535$

ii)	(3 points) Estimate the probabilities of winning each possible dollar amount for each drop location. 

```{r}
pred2 <- data.frame(drop = 1:5)
pred2 <- cbind(pred2, predict(mod.fit2, 
                     newdata = data.frame(drop = 1:5), 
                     type = "probs"))
pred2 %>% knitr::kable(digits = 3, 
                       col.names = pihat_labels, escape = FALSE)
```


iii)	
(5 points) 
Plot the estimated model. 
Make sure to include a legend to differentiate among the different lines on the plot. 
Interpret the plot. 

```{r}
# plot estimated probabilities from proportional odds model
pred2 %>% pivot_longer(cols = 2:6, names_to = "win", values_to = "probability") %>%
  ggplot() +
  geom_line(aes(x = drop, y = probability, col = win)) +
  labs(title = "Estimated probability for each win type",
       y = expression(hat(pi[j]))) +
  theme_bw()
```
    
The probabilities of winning \$100 or \$500 is a decreasing function of drop position whereas the probabilities
of winning \$0 or \$10,000 increase as drop position is increased, i.e., the drop occurs closer to the
center of the board. The probability of winning \$1,000 is a concave down function that is highest at drop position 3. The probability of winning $10,000 is highest directly when dropped directly above that
position (drop = 5), but this is also where the probability of winning nothing is highest.
    
iv)	(5 points) Estimate odds ratios corresponding to the drop variable and calculate the corresponding 95% confidence interval. Interpret the interval. 

```{r}
# OR for 1 unit increase in drop position
OR_ord <- exp(-coef(mod.fit2))
round(OR_ord, 2)
# confidence interval
CI_OR_ord <- rev(exp(-confint(mod.fit2, level = 0.95)))
round(CI_OR_ord, 2)
```

The estimated odds of the win category being below a particular level changes by 0.75 times for a 1 unit increase in the drop variable (i.e., dropping the coin 1 slot closer to the center of the board). We are 95% confident that the odds of the win category being below a particular level changes by between 0.54 and 1.03 for a 1 unit increase in the drop variable. These results suggest that there is not strong statistical evidence to conclude that the drop position changes the odds of achieving a win below a particular win level.

v)	(3 points) Perform a likelihood ratio test to evaluate the importance of the drop variable. Make sure to include the hypotheses in the correct symbolic form. 

$H_0: \beta_{1} = 0$

$H_A: \beta_{1} \neq 0$

```{r}
car::Anova(mod.fit2, test = "LR")
```

We fail to reject the null hypothesis that drop position and the log-odds of being below a particular win level and independent (p = 0.07866).

f)	(3 points) Compare the results for the multinomial and proportional odds regression models. 

Overall, the inferences from the multinomial and proportional odds model are consistent with each other. Likelihood ratio tests for independence for both the multinomial and proportional odds model suggest that the outcome of plinko is independent of the position that the coin is dropped. The only qualitative difference is that in the proportional odds model, the estimated probability of winning \$1,000 (level 3) was highest when the coin is dropped at position 3 whereas the estimated probability of that same event was highest at drop position 5 in the multinomial model.

g)	(3 points) If the goal is to maximize the probability of winning $10,000, where should the chip be dropped? Explain. 

In both models we failed to reject the null hypothesis that drop position is independent of win outcome, which means there is not statistically compelling evidence that drop position has an effect on the Plinko outcome. However, the estimated probability of winning $10,000 is highest when the coin is dropped directly over position 5 in both models, so it would be worthwhile to drop the coin in position 5 since even if the effect of drop position is not statistically significant, it may have practical significance to a contestant.

```{r echo = FALSE, eval = FALSE}
# what is the expected reward for each drop position?
reward <- matrix(c(100, 500, 1000, 0, 10000), nrow = 5, ncol = 1)
# multinomial model
as.matrix(pred1[,-1]) %*% reward
# proportional odds model
as.matrix(pred2[,-1]) %*% reward
```