---
title: "Project 1"
author: "David Nguyen"
date: "February 2, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(binom)
```

# 1. COVID tests (23 total points)

There are three types of COVID-19 tests: 1) RT-PCR, 2) antigen, and 3) antibody. Each type needs to have high sensitivity (probability test gives a positive result given individual is truly positive). The RT-PCR type is considered to have the highest sensitivity, but it also takes the longest amount of time to perform. The antigen test is used as an alternative because it takes much less time. Unfortunately, this test can be less accurate than the RT-PCR. The purpose of this problem is to examine the sensitivity of the antigen test.

The BD Veritor System is an antigen test that received an Emergency Use Authorization (EUA) from the Food and Drug Administration (FDA) on July 2. The clinical performance of the test is summarized at https://www.fda.gov/media/139755/download (see p. 12). The stated “sensitivity” is based on the correct identification of 26 positive specimens out of 31 known positive specimens. Using this data, complete the following. 

i)	(2 points) Define $\hat\pi$  as the estimated sensitivity. Compute this value.  

```{r}
w <- 26
n <- 31
sensitivity <- w/n
```

$\hat\pi$ = `r round(sensitivity, 2)`

ii)	(5 points) Calculate a 95% confidence interval for the sensitivity . While normally only one interval would be calculated in practice, calculate the interval using the Wilson, Agresti-Coull, and Clopper-Pearson expressions. Compare the intervals.  

```{r}
binom.confint(x = w, n = n, conf.level = 0.95, methods = c("wilson", "ac", "exact")) %>% # Clopper-Peason is "exact"
  mutate(width = upper - lower) %>%
  knitr::kable(digits = 3)
```

The confidence intervals for all methods are wider than 0.25, which indicates that the sensitivity of the BD veritor test is quite uncertain. The Wilson interval is narrowest, then the Agresti-Coull, and the Clopper-Pearson (CP) is widest. It makes sense that the CP interval is widest since it is *always* $\geq 1-\alpha$ (i.e., meets or exceeds the nominal confidence level), whereas the Wilson and Agresti-Coull can have true coverage probabilities lower than the nominal coverage especially when the $\hat \pi$ is near 0 or 1 (0.84 in this case).

iii)	(2 points) Interpret the Wilson interval from ii) in the context of the test. 

We are 95 % confident that the sensitivity, the probability that a sample that tested positive using RT-PCR will test positive using BD Veritor, is between 0.67 - 0.93.

iv)	(3 points) The confidence interval for the sensitivity is quite wide. Suppose a larger sample size is taken and $\hat \pi$ remains the same to two decimal places. What is approximately the smallest sample size that would result in a Wilson interval no wider than 0.05? You may use a trial and error method to find this sample size.    

```{r}
# function to compute the wilson interval at a desired width 
# will return ~ 0 when at desired width
wilson_width <- function(n, p, alpha = 0.05, width = 0.05) {
  # compute corrected probability of success
  w <- n*p
  pi.hat <- w/n
  p.tilde <- (w + qnorm(p = 1-alpha/2)^2 /2) / (n+qnorm(1-alpha/2)^2)
  # compute lower and upper wilson
  lower.wilson <- p.tilde - qnorm(p = 1-alpha/2) * sqrt(n) / 
    (n+qnorm(1-alpha/2)^2) * sqrt(pi.hat*(1-pi.hat) + qnorm(1-alpha/2)^2/(4*n))
  upper.wilson <- p.tilde + qnorm(p = 1-alpha/2) * sqrt(n) / 
    (n+qnorm(1-alpha/2)^2) * sqrt(pi.hat*(1-pi.hat) + qnorm(1-alpha/2)^2/(4*n))
  # get width of interval - desired width
  return(upper.wilson - lower.wilson - width)
}

# find minimum sample size to get desired width
n_minimum <- uniroot(wilson_width, c(31, 1000), p = sensitivity)$root %>% ceiling()

# get confidence interval with n_minimum
binom.confint(x = sensitivity*n_minimum, n = n_minimum, methods = "wilson") %>% 
  mutate(width = upper - lower) %>%
  knitr::kable(digits = 3)
```

The minimum necessary sample size to get a Wilson confidence interval of width 0.05 is `r n_minimum`

v)	(3 points) Examine the interval given for the positive percent agreement (PPA) measure on p. 12 (Table 1) of the download from the FDA. BD uses this value as the sensitivity of the test (see their definition of PPA and Table 2). What confidence level and interval did BD use in their calculations? If you cannot decide on only one interval, list all intervals that it could be. 



# 2. 