---
title: "Section 1.1"
author: "David Nguyen"
date: "February 2, 2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(binom)
```

The following reproduces some parts of STAT 875 lecture notes for section 1.1

# The binomial distribution

The PMF of the Bernoulli distribution is $P(Y=y) = \pi^y(1-\pi)^{1-y}$ for $y=0$ or $1$.

The binomial distribution is the sum of $n$ independent Bernoulli trials ($W = \sum_{i=1}^n Y_i$) and has PMF:

\begin{align}
P(W=w) & =\frac{n!}{w!(n-w)!}  \pi^w(1-\pi)^{1-w} \\
       & = {n \choose w}  \pi^w(1-\pi)^{1-w}
\end{align}

* $W$ is a random variable representing the number of successes out of $n$ trials
* $W = 0, 1, \ldots, n$
* $n$ is a fixed constant
* $\pi \in [0,1]$ is the probability of success 

Mean $E(W) = n\pi$ 

Variance $Var(W) = n\pi(1-\pi)$

# Example - field goals
Suppose a field goal kicker attempts 5 field goals during a game and each field goal has the same probability of being successful (the kick is made). Also, assume each field goal is attempted under similar conditions; i.e., distance, weather, surface,…. 

Suppose $n = 5$ and $\pi = 0.6$

What are the probabilities of getting $0, 1, 2, \ldots, or 5$ goals?

```{r}
pmf_goals <- data.frame(goals = 0:5,
          probability = dbinom(x = 0:5, size = 5, prob = 0.6))

pmf_goals %>%
  knitr::kable(digits = 2)

pmf_goals %>%
  ggplot(aes(x = goals, y = probability)) +
  geom_point(size = 2) + geom_segment(aes(xend = goals, yend = 0))  +  
  ylim(0, 0.4) +
  labs(title = expression(paste("PMF for Binomial(", pi, " = 0.6, n = 5)")), 
       y = "P(W = w)",
       x = "goals (w)")
```

#